---
title: "Anti-depressant prescriptions"
author: "L.Barnett"
date: "25/03/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

# Trend of anti-depressant prescriptions in the Liverpool population


## Methods
Using data from Liverpool CCG we modelled the monthly rate of anti-depressant prescriptions in the Liverpool population from January 2018 to February 2021, covering the pre and post-COVID pandemic time periods.
The monthly prescription rate was calculated overall by identifying how many patients had received one or more prescriptions of anti-depressants per month, and dividing by the total number of patients who were alive and contributing data in the same month. To get the rate in person-months, the crude rate was multiplied by 10,000.  
The crude rate was then stratified by; age in the time period(categorised into 10-year increments); gender; ethnicity; local IMD 2019 quintile; and shielded status.  
In order to investigate the associations between patient demographics and prescription rate of anti-depressants, logistic regression was fitted to the data.  



## Results  
Overall, the monthly rate of anti-depressant prescriptions appeared to increase over the time period.


```{r pressure, echo=FALSE, message = FALSE}
library(ggplot2)
library(readxl)
library(dplyr)
load("final analysis data.RData")
analysis %>%
  group_by(timestamp) %>%
  summarise(n = sum(antidep)) %>%
  mutate(prev = (n/494546)*10000) %>%
  mutate(timestamp = as.character(timestamp)) %>%
  ggplot(aes(x = timestamp, y = prev)) +
  geom_point(size = 2) +
  labs(x = "Year/month", y = "Prescrption rate per 10,000 P-M") +
  ylim(0, 700) +
  theme(axis.text.x = element_text(angle = 90))
```

 
Females had a higher prescription rate than males throughout the time period.  


```{r, echo = FALSE}
denom.gend <- analysis %>%
  distinct(linkpseudo, .keep_all = TRUE) %>%
  group_by(gender.y) %>%
  count()
analysis %>%
  group_by(timestamp, gender.y) %>%
  summarise(num = sum(antidep)) %>%
  left_join(denom.gend, by = "gender.y") %>%
  mutate(prev = (num/n)*10000) %>%
  filter(gender.y %in% c("Male", "Female")) %>%
  mutate(timestamp = as.character(timestamp)) %>%
  ggplot(aes(x = timestamp, y = prev,
                   color = gender.y)) +
  geom_point(size = 2) +
  labs(x = "Year/month", y = "Prescrption rate per 10,000 P-M") +
  ylim(0, 1000) +
  theme(axis.text.x = element_text(angle = 90))
```


Patients between the ages of 50 and 59 years had the highest prescription rate at each time point, followed by patients aged 40-49 years. The age group with the lowest prescription rate were 0-19 years old, followed by those aged 70 years and older.


``` {r, echo = FALSE}
denom.age <- analysis %>%
  distinct(linkpseudo, .keep_all = TRUE) %>%
  group_by(age.group.time) %>%
  count()
analysis %>%
  group_by(timestamp, age.group.time) %>%
  summarise(num = sum(antidep)) %>%
  left_join(denom.age, by = "age.group.time") %>%
  mutate(prev = (num/n)*10000) %>%
  mutate(timestamp = as.character(timestamp)) %>%
  ggplot(aes(x = timestamp, y = prev,
             color = age.group.time)) +
  geom_point(size = 2) +
  labs(x = "Year/month", y = "Prescrption rate per 10,000 P-M") +
  ylim(0, 1200) +
  theme(axis.text.x = element_text(angle = 90))
```


Patients living in the most deprived(?) areas of Liverpool had the highest rate of anti-depressant prescriptions, whilst those in the least deprived had the lowest.


``` {r, echo=FALSE}
denom.dep <- analysis %>%
  distinct(linkpseudo, .keep_all = TRUE) %>%
  group_by(local_imd_quintile.y) %>%
  count()
analysis %>%
  group_by(timestamp, local_imd_quintile.y) %>%
  summarise(num = sum(antidep)) %>%
  left_join(denom.dep, by = "local_imd_quintile.y") %>%
  mutate(prev = (num/n)*10000) %>%
  mutate(timestamp = as.character(timestamp)) %>%
  ggplot(aes(x = timestamp, y = prev,
             color = local_imd_quintile.y)) +
  geom_point(size = 2) +
  labs(x = "Year/month", y = "Prescrption rate per 10,000 P-M") +
  ylim(0, 1000) +
  theme(axis.text.x = element_text(angle = 90))+
  scale_color_discrete(labels = c("Most deprived", "2", "3", "4", 
                                  "Least deprived", "missing"))

```


After stratification by CEV status, patients who did not shield had higher rates of anti-depressant prescriptions than those who did shield.


``` {r, echo = FALSE}
denom.shield <- analysis %>%
  distinct(linkpseudo, .keep_all = TRUE) %>%
  group_by(shielded) %>%
  count()
analysis %>%
    group_by(timestamp, shielded) %>%
  summarise(n = sum(antidep)) %>%
  left_join(denom.shield, by = "shielded") %>%
  mutate(prev = (n.x/n.y)*10000) %>%
  mutate(timestamp = as.character(timestamp)) %>%
  mutate(shielded = factor(shielded, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  ggplot(aes(x = timestamp, y = prev, color = shielded)) +
  geom_point(size = 2) +
  labs(x = "Year/month", y = "Prescription rate per 10,000 P-M",
       color = "CEV") +
  ylim(0, 2000) +
  theme(axis.text.x = element_text(angle = 90))
```


Stratifying the average monthly number of anti-depressant prescriptions by gender, and COVID period found that both groups had a higher number of anti-depressants during the pandemic period.


``` {r, echo = FALSE}
analysis %>%
  mutate(covid.period = factor(covid.period, levels = c(0, 1),
                               labels = c("Pre", "During"))) %>%
  filter(gender.y != "Not specified",  gender.y != "Unknown") %>%
  group_by(timestamp, covid.period, gender.y) %>%
  summarise(n = sum(antidep)) %>%
  group_by(covid.period, gender.y) %>%
  summarise(c=n(), avg.monthly = mean(n), sd.monthly = sd(n)) %>%
  mutate(se.monthly = sd.monthly/sqrt(c)) %>%
  ggplot(aes(x = gender.y, y = avg.monthly, fill = covid.period)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = avg.monthly-(1.96*sd.monthly), ymax = avg.monthly+(1.96*sd.monthly)),
                position = position_dodge(0.9), width = 0.1) +
  labs(x = "Gender", y = "Average monthly prescriptions", 
       fill = "COVID Period")
```


After stratification by age group, the average number of anti-depressants was higher during the COVID period for all age bands.

``` {r, echo = FALSE}
analysis %>%
  mutate(covid.period = factor(covid.period, levels = c(0, 1),
                               labels = c("Pre", "During"))) %>%
  group_by(timestamp, covid.period, age.group.time) %>%
  summarise(n = sum(antidep)) %>%
  group_by(covid.period, age.group.time) %>%
  summarise(c=n(), avg.monthly = mean(n), sd.monthly = sd(n)) %>%
  mutate(se.monthly = sd.monthly/sqrt(c)) %>%
  ggplot(aes(x = age.group.time, y = avg.monthly, fill = covid.period)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = avg.monthly-(1.96*sd.monthly), ymax = avg.monthly+(1.96*sd.monthly)),
                position = position_dodge(0.9), width = 0.1) +
  labs(x = "Age group", y = "Average monthly prescriptions",
       fill = "COVID period")
```


Stratification by IMD 2019 quintile showed that the most deprived areas had an increase in monthly average anti-depressant prescriptions during the COVID-19 pandemic, similarly to all other deprivation quintiles.

``` {r, echo = FALSE, message = FALSE}
analysis %>%
  mutate(covid.period = factor(covid.period, levels = c(0, 1),
                               labels = c("Pre", "During"))) %>%
  group_by(timestamp, covid.period, local_imd_quintile.y) %>%
  summarise(n = sum(antidep)) %>%
  group_by(covid.period, local_imd_quintile.y) %>%
  summarise(c=n(), avg.monthly = mean(n), sd.monthly = sd(n)) %>%
  mutate(se.monthly = sd.monthly/sqrt(c)) %>%
  mutate(local_imd_quintile.y = factor(local_imd_quintile.y, 
                                       levels = c("1", "2", "3", "4", "5", "NULL"),
                                       labels = c("Most deprived", "2", "3", "4", 
                                                  "Least deprived", "missing"))) %>%
  
  ggplot(aes(x = local_imd_quintile.y, y = avg.monthly, fill = covid.period)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = avg.monthly-(1.96*sd.monthly), ymax = avg.monthly+(1.96*sd.monthly)),
                position = position_dodge(0.9), width = 0.1) +
  labs(x = "IMD 2019 quintile", y = "Average monthly prescriptions",
       fill = "COVID period")
```


